Quarterly Meeting Report

### 2013-12-14 pnsac/newsletter/article_Feature1.tex
Preparation for December 2013 newsletter.

\title{My Days at the Home of the Merlin Engine}
\subtitle{Part 3}
\author{Richard Lodge}

@@

``Lodge, the expenses on the charity wagon go to Miss Prod''. What was a
charity wagon and who was Miss Prod? I was working in the Overheads
Department of the Rolls-Royce Aero Engine Division in Derby. As a
young, ambitious accountant, I did not want to admit to the people in
the department that I did not know what they were talking about. I was
conscious that I had spent several weeks of initiation under the
guidance of the old timer, Les Hart, and should now know something
about the company. I was sitting in front of a huge leather bound
ledger where accounting information was recorded. Eventually, by
listening to other people talking around me, I established that a
charity wagon was a much prized company car and that Miss Prod was not
a person but referred to ``miscellaneous production'' work at R-R. Put
in plain English, the sentence meant that the cost of repairing this
individual's car was to be charged to a special account in the
miscellaneous production section of the overheads. There was so much
to learn!  I was completely bewildered. Each day was a challenge and
the thought that aero engines were produced in the surrounding
factories seemed very far away.

On being placed in the Overheads Department, I was given an internal
mail address. The company was so large that a worldwide internal mail
system had been established and I rapidly found out that I was at the
bottom of the heap. My address was JD/JS/LOD which meant that JS was
my immediate supervisor and JD was his boss. If you were important in
the company, your mail would come to you with just your initials such
as JD. Everybody with any ambition wanted to get a single initials
mailing address. If you were really close to God, such as a director
of the company, your mailing address would be Psn for the CEO, Sir
Denning Pearson, or DPH for David Huddie, the Director of Engineering.

There were continuous battles between the engine development engineers
and the accountants. When the engineers spent money, they tried
wherever possible to avoid having it charged against their
budget. Their favourite game was to charge the cost into the overheads
where it would be lost in a mass of other expenditure such as emery
paper and toilet rolls. Our task was to catch them at their game and
reallocate the expense back to them. The development engineers were
very skilled at this part of their work and became so good at it that
they were able to hide major cost overruns, sometimes
indefinitely. Little did I realize that I was helping to perpetuate
the culture that was one of the causes of R-R going bust a few years
later.

In the 1960s R-R was a paternalistic organization, steeped in
tradition, still operating with many cost-plus government contracts,
vast amounts of paper and virtually no computers. The Ministry of
Aviation would regularly send auditors into the company to check how
public money was being spent. These gentlemen had a permanent office
allocated to them and would regularly attack the expenses recorded by
the Overheads Department. When this happened, we and the development
engineers became great buddies for a short time. Our unspoken job was
to confuse the government auditors as much as possible so that they
would not report back to their masters that things were not as they
should be. We became quite good at our confusion duties. It was not a
difficult job because we were frequently confused by all the paper and
numbers as well.

R-R continued to deliver new aero engines to customers and I gradually
started to understand what I was doing. After a few months, my
supervisor was promoted and I was appointed supervisor of the
Overheads Department, much to the disgust of many of the people in the
department. Although my pleasure was to be short-lived, I was
delighted to get this promotion. By the time of my promotion, I had
been around many of the facilities of R-R both in the Derby area and
in Scotland. I felt that I knew my way around the company.

Almost everyone working in the Overheads Department, including the
three section leaders under me, were older and more experienced than
I. They did not like this young upstart accountant becoming their
boss. Underground warfare started between us. Several of the men had
fought through WW2 and were well practiced at the gentle art of ``dumb
insolence''. The three section leaders came to me one day and told me
they really disliked me being their boss. This was not good for my
ego. I suggested to them that the four of us should go out after work
and have a beer and that on this occasion I would just be one of them
and not their boss. I bought the first round of beer and they really
tore into me. After several more rounds of beer, we agreed to a truce
and decided on a new strategy. I would stop trying to micromanage them
and they would stop their dumb insolence. Of course, they never
admitted that they were practicing dumb insolence. Life became much
better after our beers. Miss Prods, charity wagons, drop-offs and all
the other R-R jargon became normal for me with the active co-operation
of my three section leaders.

Apart from the pay rise upon my promotion I had now moved up in the
company mailing system and my address had become JD/LOD. This probably
helped my head to swell until the three section leaders brought me
rapidly back to earth.

I must have learnt a few of life's lessons about people management
because when I was again promoted, my three section leaders tried to
encourage me to refuse the promotion. No chance. I was once again
moving up the mailing system and would in future have my mail
addressed to LOD, without a more senior person's initials appearing
before mine. I was going in the right direction.

>>>>

\title{My Days at the Home of the Merlin Engine}
\subtitle{Part 4}
\author{Richard Lodge}

@@

One of the most mysterious departments at Rolls-Royce was the
Technical Publications Department. Word processing was unheard of in
the 1960s and every time an engine modification was made, it resulted
in much work documenting even a small change and distributing the
updated documentation to all engine users around the world. I never
spent any time working with the people in this department even though
there was a complete building devoted to the work. Compared with the
resources allocated to engine development, the technical publication
side of the business was very small and it always remained a mystery
to me.

Since retiring from my career as an accountant in Canada, I have
become active on the actual restoration work of the North
Star. Understandably, I have a special interest in working on the
Merlin engines. Nobody now working on the restoration has ever worked
on a Merlin when it was in service. It is therefore necessary at all
times to refer to the various publications and manuals we have managed
to accumulate for the North Star.

It is only now, actually working on the engine and referring
frequently to the exploded diagrams that I realize how complicated the
work of the Technical Publications Department must have been. The men
and women who worked there were largely unsung heroes in that it was
very tedious and unglamorous work which was of great importance for
the safe and trouble-free running of engines in service. I can
remember that I would hear from time to time of the difficulties of
ensuring that publication updates were actually delivered to many
remote places in the world and the considerable logistical
complications of ensuring that the packages arrived safely and
furthermore that the updates were actually inserted in the relevant
manuals, particularly as many of the engine users did not speak
English. I cannot remember now if any of the documentation was
translated into other languages although I have a recollection that we
employed people who could translate into French and Spanish.

After a couple of years I was given a new job in the Long Range
Planning Department. I was one of three accountants working in this
office under a very senior accountant. Here we looked at the costings
of the development work, the projected sales  and production costs on
new engine designs. This was the time when the Spey was going into
service and the the company had signed a fixed price contract for the
RB211 engines for the new Lockheed 1011. This was a major project,
championed by David Huddie the Chief Engineer. Computer spreadsheets
did not exist at that time - pencil and columnar paper were the order
of the day, together with efficient erasers. Long range forecasts of
25 years required many yearly columns and each time we produced
figures that the engineers did not like, we were told to rework the
numbers until we produced an acceptable answer. We were not limited in
the number of erasers we could use. The usual way of cutting costs was
to reduce the forecast development and testing time. Of course this
did not work in real life.

We were often dealing with the directors of R-R who lived on the top
floor of the Nightingale Road offices. Many of them were engineers and
they had little time for young bean counters. It was still a time of
optimism for the company. The huge contract for the Lockheed engines
had been signed and the company was still enjoying its reputation as
the company whose Merlin engines had saved Britain in the air
war. This was in 1965/6 and it was several years before our
projections of doom resulted in the bankruptcy of the
company. Fortunately I left the company in 1966 and was not involved
in what must have been a horrible process as a symbol of British
engineering excellence slowly went to the wall.

After the chaos caused by the demise of Rolls-Royce Limited, the
British government formed a new nationalized company to take over the
remains of the old one. This has been very much a success story and
the newly privatized Rolls-Royce is now a highly profitable and
innovative company. I look back on my days at R-R with nostalgia but I
now realize that much of the time we were all living in a fool's
paradise.

### 2013-12-14 pnsac/newsletter/article_Feature2.tex
Removed text from article_Feature2.tex to match wrtier's original content.
Preparation for December 2013 newsletter.

\title{Building troop seats for North Star 17515}

\author{Rolf Geiger}

@@

There are no original parts or drawings for the seats. Fortunately,
the tubing and brackets that connect the seats to the aircraft wall
were still in place in the aircraft. Also, two lower tubing support
brackets were found hiding behind the galley, allowing us to produce
exact copies of the original. Everything else concerning the seats is
missing and needs to  be replaced.

\end{multicols}

%\begin{figure*}[htbp]
\begin{figure*}[h]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{ass.eps}
   %caption of the figure 
   \caption*{\small \em Troop seat construction detail.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:ass}
\end{figure*}

\begin{multicols}{2}

To find the size and the number of seats to be produced we measured
the space between the anchoring studs in the aircraft floor, which the
legs snap onto. This gave us the width of the seats. Thus, we will
produce (5) three-seaters plus (6) four-seaters, providing seating for
39 troops. No litters are planned.

Mike Irvin (our Museum coordinator) decided to use seats from a
helicopter in the Museum collection as an example for our seats. We
are more or less copying the parts of these seats, considering that we
have to adapt our methods to the equipment available to us.

To start the project, drawings of the parts were made. The list of
materials of these parts calls mainly for aluminum bar stock, aluminum
tubing, steel tubing, aluminum sheet and steel sheet.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{rolf.eps}
   %caption of the figure 
   \caption*{\small \em Rolf Geiger at work on the troop seats.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:rolf}
\end{figure}

Thus, we need to produce (15) different metal parts, including: lower
tubing plus support brackets and clamps, spreader bars with two shell
clamps each to clamp onto the lower support tubing plus the front
brackets holding the spreader bars to the frontal support tubes, leg
brackets connecting the legs to the frontal support tubes plus
secondary brackets (slipping over the leg brackets) holding a spring
loaded pin with which the seat fabric will be tautened, legs,
including the lower clamping mechanism fastening the legs to the
floor. Most parts need (22) of each, except the shell clamps for the
spreader bars (44), hooks (117), leg grippers (44).

The procedure in the production of these parts is as with any project:
Take the finished product and then work backwards to determine the
production stages.

In the case of a formed sheet metal part a drawing of a flattened
blank is produced by calculating bend radii and adding that to the
straight surfaces (in effect un-bending or stretching the finished
part).

Experimental operations may be needed to find the shape and dimensions
of the blanks.  Tooling may be needed to produce the formed sheet
metal parts.  The blanks are made by shearing and sawing, according
to the parts drawing.

In the case of a machined part, it is usually produced from bar stock:
Parts are cut to length and subsequently machined as per
drawing. Partial in-between drawings may be necessary to produce the
parts in progressive stages.

To avoid the higher costs associated with surface plating the metal
parts, (as per original parts) we chose painting them instead, which
will be acceptable given the limited use of the finished product.

>>>>

\title{RCAF Overseas Ferry Operations}
\author{AJS (Tim) Timmins, RAD/NAV/AI}

@@

Over a three year period starting in 1952, with Operation Leap Frog,
the deployment of 12 Sabre squadrons from Canada to Air Division in
Europe, No 1 Overseas Ferry Unit (1OFU), based at St Hubert, Quebec,
delivered over 800 aircraft. 426 Squadron North Stars flew support
missions, carrying servicing personnel and providing enroute
navigation assistance, airborne radio beacons known as duck
butts. Besides the Sabre squadrons, four CF100 Squadrons were
delivered to Air Division and 53 CF100 aircraft to the Belgium Air
Force. Silver Stars and Expeditors were flown to Europe for delivery
to NATO members under the Mutual Aid agreement. My personal experience
with ferry operations was with Beechflight, the delivery of 25 C45s
(Expeditors) to France and Portugal.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{rcaf_c45.eps}
   %caption of the figure 
   \caption*{\small \em RCAF C45 Expeditor a military variant of the Beech 18.Purchased in large numbers during WW II and after for pilot training, navigator training, and as a utility transport.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:rcaf_c45}
\end{figure}

Beachflight was a Training Command operation, under the command of
Wing Commander Harry Forbell. Twenty six expeditors were assembled at
Trenton where they were fitted with a 100 gallon fuel tank in the main
cabin, plus safety and navigation gear required for the planned
route. A pilot and navigator were assigned to each aircraft. I flew
with Squadron Leader Brian, Marfleet, the designated Deputy Operation
Commander. The aircraft were to fly in formation, four sections of
five and one of six, within block airspace.

On 20 May 1959 Operation Beachflight was launched. Sections departed
Trenton for Goose Bay at designated intervals in order to maintain
separation along the route. Big trouble, the weather within the
assigned block airspace deteriorated to the point where aircraft could
not maintain contact with each other and abandoned their
formations. Now it was everybody for himself. The 26 Expeditors were
up, down, right, left, within the block airspace and no doubt outside
it. My section dispersed and we did not see any of them until we
arrived at Goose Bay. Not a great start to the operation, but everyone
arrived safely at Goose Bay, so a success in that sense. Our flight
time was 6:45.

The flight from Goose Bay to Frobisher (Iqualuit) the next day was
uneventful. Our Flight time was 6:05.

As I recall, we had a weather delay in Frobisher and did not depart
until 24 May. The plan was for each section to proceed to Sondrestrom
AFB (Kangerlussuag-big Fjord) located on the West coat of Greenland,
refuel and carry on over the icecap to Keflavik. The Deputy Commander
would lead the 5th and last section. Our aircraft was unserviceable so
we had to take the spare, the 26th aircraft. The unserviceable
aircraft was repaired and returned to Trenton. We were late into
Sondrestrom, to find that section four was still on the ground because
the three earlier sections had experienced heavy icing over the ice
cap.  It was not known if the crews and aircraft were safe, a great
concern as the Expeditor was a notoriously poor performer in icing
conditions. We assembled in the bar to await word on the fate of our
mates. All the aircraft made it over the ice cap and arrived safely in
Keflavik, which was ample reason for us to celebrate another
success. Our flight time to Sondrestrom was 3:50.

We departed Sondrestrom on 27 May after a two day delay waiting for
improved weather over the ice cap. The flight to Keflavik was
uneventful. Our flight time was 5:10.

The first three sections waited for us in Keflavik. Every crew had a
story to tell about their experience over the ice cap. Fifteen
aircraft were loaded with ice and unable to maintain altitude. Their
only salvation was to make it past the ice cap and out over the water
where they could descend to warmer air and shed the ice. Thankfully,
all of them did.

The east coast of Greenland is very rugged with numerous high
peaks. Several crews in their descent out of the icing conditions
broke cloud between these peaks. These stories were oft repeated and
all called for another round of drinks.

The next day, 28 May, all the sections proceeded to Prestwick. The
weather was perfect and I got a lot of stick time because my pilot was
very tired after the debriefings in Keflavik. Our flight time was
5:10.

We remained overnight in Prestwick and flew to 1 Wing, Marville,
arriving mid afternoon on 29 May.  The aircraft were prepared for
delivery to the French and Portugese air forces. The cabin fuel tanks
were removed as well as all the safety and navigation gear.

On 3 June we delivered 19 Expeditors to the French Air Force at
Chateaudun, located not far from Chartres. The flight crews were
picked up by C47 Dakotas and returned to Marville. Nineteen crews
boarded a 426 Squadron North Star for return to Trenton. Of course the
Commander and his Deputy remained behind to deliver six expeditors to
the Portugese Air Force in Lisbon.

We departed Marville on 4 June for Bordeaux where we planned to refuel
and then proceed along an air route over the Pyrenees Mountains to
Lisbon. The weather along the planned route was very bad with
intensive thunder storms over the mountains. Ever resourceful, our
Commander decided to file a flight plan west over the Bay of Biscay,
clear of Spain, then south to Lisbon. This would be a long overwater
flight and we had no extra fuel, safety equipment or nav gear. We took
off and were in the process of forming up in two sections of three,
when several light military aircraft flew through our formation. No
one saw them coming, it happened so fast there was no reaction. Not a
very good start to the day.

Our Commander led us west over the Bay of Biscay, maintaining VFR
under a cloud layer. The cloud  base got lower and lower, we were down
with the shipping and it was decided to climb above the cloud, VFR on
top. This worked for a while but the cloud tops kept creeping up over
10,000 feet and we had no oxygen. The Commander decided to descend and
try for better conditions. The Deputy Commander, my pilot, decided to
stay high; our six plane formation was split three high and three
low. I was now doing the navigation for the high section. With nothing
but a chart to work with, I managed to make use of the marine beacons
along the coast to guide us west then south to Lisbon. We arrived at
Lisbon before the low section and the Deputy Commander wisely elected
to hold until the Commander arrived and landed. Our flight time was
5:45.

The next day we participated in a small handing over ceremony. Mission
accomplished!

### 2013-12-14 pnsac/newsletter/article_ProgReport.tex
Newsletter: final changes incorporated.
Preparation for December 2013 newsletter.

\title{Project Manager's Progress Report}
\subtitle{October 2013}
\author{Bruce Gemmill}

@@

Engine 3 is almost completely reassembled, with only the supercharger
and intercooler preheat assemblies left to install on the engine,
which was moved to the QEC engine frame in early summer.  The work has
slowed somewhat, due to summer down time for staff and volunteers, but
we continue to make steady progress.

The auxiliary gearbox was recently removed from the engine firewall,
and is now being disassembled.  This is the last major assembly that
must be overhauled before the engine can be installed on the aircraft.

\section{Engine Frame}
\label{engineframe}

Now that the engine and frame are joined, the major work still
required is rebuilding the many cowl panels and the radiator flaps
that surround the engine.  There is a lot of corrosion on the pieces,
and each must be disassembled, cleaned, then repaired before painting
and reassembly.  This work must be done carefully to ensure the cowl
panels fit properly when installed on the QEC.

This work has gone slowly due to the small number of volunteers who
are able to do this type of repair.

@@

The new galley has now been installed in the crew lounge, and work
overhead in the ceiling is also complete.  This involved installing
new insulation and the Janitrol heaters, along with refurbished
heating ducts, and some new fuel lines.  The ceiling panels were also
installed.

The doors on the main electrical panel were restored and installed.
New circuit legend cards were produced to indicate the circuit breaker
and junction box connections.  These were laminated and installed
inside the electrical panels.  Some additional investigation is needed
to complete the electrical circuit connection diagrams that were badly
faded.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{galley_interior.eps}
   %caption of the figure 
   \caption*{\small \em New galley -- interior view.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:galley_interior}
\end{figure}

One section of floor was patched after removing a corroded area.  One
additional repair needs to be done under the floor of the crew lounge.
Until this is complete, the new cushion flooring cannot be
installed. This has also delayed the installation of all of the crew
lounge accessories, such as the table, seats and secure storage bin,
all of which are complete but waiting in storage.

@@

The forward and rear belly compartments are now complete.  The two
battery elevators that were removed have been restored and installed
behind the nose wheel.  The battery compartment covers were repaired
but need painting and stencilling.

The firewall on nacelle 3 is being cleaned in preparation for the
installation of the engine sometime this winter.

Further progress has been made on the set of troop seats being
fabricated for the main cabin.

@@

Over the next year, we hope to have engine Nr 3 completed and engine
Nr 4 removed.  We will complete the reassembly of the crew lounge and
galley.  We then plan to work on the main cargo compartment, including
refurbishing the main heater duct and other fittings, and removing
floors to begin cleaning and repairs under the cargo floor.  We may
yet get to work on the four engine nacelles.

>>>>

\title{Project Manager's Progress Report}
\subtitle{December 2013}
\author{Bruce Gemmill}

@@

Engine 3 is nearing completion.  The supercharger and intercooler
preheat assemblies have been installed, along with the propeller
reduction gearbox.  While this work was being carried out by the
engine crew, the remainder of the volunteer work force busied
themselves completing the numerous cowl panels, pipes and hoses and
other pieces needed to complete the engine.  Two items that seemed to
take forever were the large steel exhaust shrouds.  A lot of hammering
was needed to pound out years of dents, then numerous rivets and all
the Dzus fasteners and springs needed to be replaced.  The complex
shape of the exhaust shroud made this challenging work.  Further
delaying completion was the search for suitable clear coat to protect
the steel from corrosion.  The old clearcoat we used successfully on
engines 1 and 2 is no longer available.  Mike Irvin finally located a
suitable replacement, and the shrouds were clear coated and serial
numbers were stencilled on.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
   %caption of the figure 
   \caption*{\small \em Garry installing the intercooler on engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:Eng3Intercooler}
\end{figure}

The auxiliary gearbox was disassembled, and significant corrosion was
found on the bearings inside.  Bearings had to be ordered from
England.  Meanwhile, the gearbox was cleaned and painted.  At the same
time, the large DC generator was dismantled, repainted and
reassembled, as well as the air pump and the tachometer generator and
propeller synchronizer.  Once the new bearings were installed, the
gearbox was completed and the generator and air pump were attached.
Finally, the gearbox was installed on the firewall, which is now
finished, awaiting the installation of engine \#3.

@@

Work has slowed on the forward section of the aircraft.  The repair
needed under the floor of the crew lounge is on hold until a suitable
piece of extruded aluminum can be found.  Until this is complete, the
new cushion flooring cannot be installed. This has also delayed the
installation of all of the crew lounge accessories, such as the table,
seats and secure storage bin, all of which are complete but waiting in
storage.

@@

The forward and rear belly compartments are now complete.  All the
frame pieces and hardware needed for the troop seats have been made,
and work is progressing on the seats and backs.  One section has
undergone a trial fit, and all pieces seem to fit perfectly.

@@

The museum has advised all North Star volunteers that our work must be
put on hold for several months while a major exhibit for the 100th
anniversary of the beginning of the First World War is prepared.  Our
work space in the restoration hangar is needed to assemble and prepare
several aircraft that will be added to the display space in March.
Until then our work on the aircraft is suspended, but we don't expect
this delay to have a serious impact on the overall project.  We look
forward to resuming work in the Spring.

### 2013-12-14 pnsac/newsletter/article_SpecialEvents.tex
Preparation for December 2013 newsletter.

\title{Recent Special Events}
%\subtitle{Visit to Nav Canada Montreal A.C.C. and Bombardier}
\author{Bill Tate}

@@

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=1.5]{loch_march.eps}
%    %caption of the figure 
%    %\caption*{\small \em North Star engine 3.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:logo}
% \end{figure}

\section{Loch March Golf Day}
\label{golf}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=1.5]{loch_march.eps}
   %caption of the figure 
   %\caption*{\small \em North Star engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:logo}
\end{figure}

The PNSAC 2013 Annual Golf Day at Loch March started with breakfast in
the Club House, 14 members and friends in attendance.

Prior to breakfast, start times were arranged in such a way that there
was a mixing of participants, so that those who were not aware of what
PNSAC does had the opportunity to talk to members involved in the
project.

The first group started out in light rain showers and thirty minutes
later the skies cleared for a bright sunny day. With the extra
rainfall this year the course was found to be in a very lush
condition, and as usual for Loch March the greens were very fast.

Danielle Nadon the Club Professional graciously gave PNSAC two 18 hole
certificates as prizes, one for the closest to the pin and the other
for the longest drive, which were both won by Gary Hall. He also came
in with low gross for the day at 79. Our own Ron Lemieux was paired
with Gary and in Ron's own words ``I had nothing to do with the score;
I was riding on Gary's coat tails''.

Loch March is a very wonderful facility for golf and has an excellent
club house as well. The course can be challenging, but there is
something for everyone who plays golf. Even if you do not golf, it is
well worth the drive to enjoy the patio. For details please visit the
web site: {\normalfont\color{blue}\texttt{\url{http://www.lochmarch.com}}}

This year we had another prize which was given to us by New York New
York Hair Salon, - a gift certificate of \$100:00 for hair
services. So a new category of winner had to be chosen, which was for
the best dressed in our distinctive PNSAC clothing. To ensure fairness
in the award I had Alfredo Porco (who has been my barber for 30 years)
make the award. It went to our "special seamstress" Karen Lochhead.  I
should mention there are now three generations of Porcos at New York
New York, - Alfredo, his son Vince and Vince's daughter Marina.

For information on New York New York please visit their web site:
 {\normalfont\color{blue}\texttt{\url{http://www.nynyhairsalon.com}}}

Looking forward to our fourth golf day next year, we will work on more
prizes for our members and guests in 2014.

\section{Trip to the Montreal Area Control Centre and Bombardier Aerospace}
\label{yultrip}

On Friday October 4th, 33 members and guests departed the Canada
Aviation Space Museum (CASM) for another planned special event. First
off, special thanks to Bryon Mask (retired Air Canada Captain), one of
our members who assisted me with contact information at Nav Canada
which was required to have our visit approved. Bryon is a highly
respected individual in Flight Safety and Accident Investigation in
Canada.

We were met and welcomed by Richard Snider, Duty Manager at the
Montreal ACC. We visited the Training Centre where training is
provided to new Air Traffic Controllers and required recurrent
training is offered. We were also shown the tower simulator which has
a day/night 360 degree display where various airport scenarios can be
brought up to assist in training. My personal favourite was the
display showing a landing as seen from the cockpit, which even showed
it as a 360 degree display (first time I have experienced the rear
view mirror look for a landing).

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca3.eps}
   %caption of the figure 
   \caption*{\small \em In the 360 degree tower simulator.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca3}
\end{figure}

Our next stop was the Technical Operations Centre where staff monitor
all radar and computers. ACC has dual systems with one system picking
up automatically in case the other system fails. Communications has a
dual back-up digital and analogue system.  Communications to aircraft
are by conventional VHF with some interesting back ups such as
Controller Pilot Data Link Communications (CPDLC). The easiest way of
explaining this is Text Messaging Sat Com in which either the aircraft
can call the phone number of the ACC or the ACC can call the dedicated
number for the aircraft (\$6:00 a minute cost) and as a back up by HF
radio through a Flight Service Station.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca2.eps}
   %caption of the figure 
   \caption*{\small \em Air traffic controllers at work.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca2}
\end{figure}

In the Operations Room, we could hear some controllers talking very
quietly. Radar displays offer a great deal of information to the
Controller, which also helps keep talking to a minimum. A sense of
humour prevailed at the work station for Northern Quebec up to
Iqaluit, where a Motel 6 logo was mounted over the work station.

In total we were on site for close to 2 1/2 hours which I found to be
a fantastic experience, seeing the ATC environment from the other
side.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca1.eps}
   %caption of the figure 
   \caption*{\small \em The Operations Room.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca1}
\end{figure}

With kind assistance in the form of a recommendation from Mae, my YUL
expert, we broke off for lunch at Le D\'{e}jeuner Cosmopolitain which
handled our group in a very friendly efficient manner. I should note,
this is the spot where Air Canada retirees meet for their monthly
luncheon dates.

After lunch we proceeded to the Bombardier Production plant where we
were met by three very knowledgeable tour guides. They were Alex
Ritchie, Fernand Richer, and Cristian Racine who together gave a very
informed hour and half tour of the production line. The tour was
divided into two areas, assembly and finishing. At the beginning of
the assembly we saw where the fuselage and wings are mated and
proceeded on to where systems are tested such as hydraulic systems,
then to the finishing area where interiors are installed. It was
something to see how clean and bright this plant is and how the many
sub assemblies come together. The tail assembly built in China comes
to the production line and meets with part of the fuselage built in
Mexico, just in time for the assembly.

We ended our visit in the delivery room where customers pick up their
aircraft all ready to go, all yours for approximately \$30 million
dollars.

We enjoyed an excellent dinner at the Auberge Willow Inn prior to our
return to the CASM at the end of a very enjoyable 14 1/2 hour day
together.

>>>>

\title{Quarterly Meeting Report}
\author{Drew Hodge}

@@

On Saturday 7th December 2013, Project North Star Association of
Canada (PNSAC) held a quarterly meeting in the auditorium of the
Canadian Aviation and Space Museum (CASM) in Ottawa. It was a
memorable meeting with over 200 people in attendance. Those in the
audience that day were able to hear the story of the Gimli Glider told
in person by Captain (retd.) Robert (Bob) Pearson, the pilot of Air
Canada Boeing 767-200, fin number 604, when it made an unscheduled
stop at Gimli Manitoba, thirty years ago.

The meeting started with an introduction by Bill Tate, Project North
Star's Vice President.  Bill began by welcoming the audience, which,
as well as PNSAC volunteers, included members of Vintage Wings of
Canada, Ottawa Airport Watch, the Canadian Aviation Historical
Society, the College of Professional Pilots of Canada, CASM volunteers
and staff, Air Canada and WestJet pilots, and visitors from the
general public, one of whom had travelled to Ottawa from Vancouver for
the meeting. After describing the goals of Project North Star and
asking PNSAC volunteers to stand and be recognized, Bill invited the
audience to visit the world class exhibits in the Museum after the
meeting.

Bill's acquaintance with Bob Pearson goes back thirty-five years, when
Bill was Captain Pearson's Second Officer on Boeing 727s based in
Montreal. His duties in the cockpit back then gave Bill ample
opportunities to watch Captain Pearson at work. Bill described Captain
Pearson as a "highly professional pilot" who, "wearing his command
lightly", demonstrated "highly polished interpersonal skills and a
delightful sense of humour".  He then invited Captain Pearson to tell
the story of flying unscheduled into Gimli.

Captain Pearson spoke for a little over an hour with no notes and with
reference to a few appropriate slides.  He described eloquently how,
because of errors in the conversion of metric measurements, his Boeing
767-200 ran out of fuel and lost both engines over Manitoba on the way
from Ottawa to Edmonton. The full story is well documented in the book
"Freefall: From 41,000 feet to zero - a true story, William and
Marilyn Hoffer, Simon \& Schuster, 1989", but to hear it told in
person by the pilot in command that Saturday, 23rd July, 1983, was a
rare privilege.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{Pearson004.eps}
   %caption of the figure 
   \caption*{\small \em Captain (retd.) Robert Pearson.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:pearson004}
\end{figure}

The port engine stopped first at 41,000 feet prompting the decision to
divert to Winnipeg.  At 28,000 feet the starboard engine ran down too,
and Maurice Quintal, Captain Pearson's First Officer, made
calculations that showed they couldn't reach Winnipeg.  Quintal
advised Captain Pearson to try for the nearest runway on the former
RCAF aerodrome at Gimili, twelve miles away.  Some eight or ten
minutes after losing the first engine, Captain Pearson side-slipped
the big Boeing down to an old runway that was being used as a racing
track by members of the Winnipeg Sports Car Club. The nose wheel
collapsed on landing, but the aircraft came to a stop and no one was
seriously injured. The racers helped to put out a minor fire in the
nose.

Captain Pearson talked about the inquiry that followed the incident
and its eventual exoneration of both pilots.  The aircraft was flown
to Winnipeg just two days after the Gimli landing and saw service with
Air Canada for another twenty-five years.  When the Boeing was retired
to the Mohave desert in 2008, Captain Pearson was at the controls to
make that last landing in Arizona.

Following his talk, Captain Pearson answered questions from the
audience.  Richard Lodge, PNSAC President, thanked Captain Pearson and
presented him with a PNSAC golf shirt, a baseball cap, a coffee mug,
and finally a certificate conferring upon him an Honorary Life
Membership of the Project North Star Association.  Captain Pearson
then graciously agreed to make presentations of volunteer-hours
certificates and 10th Anniversary photos to PNSAC volunteers.  The
following volunteers received certificates for the hours shown:

\begin{itemize}
  \item Bruce Gemmill (7000 hrs)
  \item Ted Devey (4000 hrs)
  \item Bill Tate (3000 hrs)
  \item Richard Lodge (1000 hrs)
  \item Bruce Grant (1000 hrs)
  \item Robert D\'{e}sjardins (1000 hrs)
  \item Peter Trowbridge (1000 hrs)
  \item Charles Baril (1000 hrs)
\end{itemize}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{Pearson007.eps}
   %caption of the figure 
   \caption*{\small \em Richard Lodge presenting Captain Pearon with a
     certificate for Honorary Life Membership in PNSAC.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:engine3propcowl}
\end{figure}

To conclude the meeting, Captain Pearson presented signed copies of
the book "Freefall: From 41,000 feet to zero - a true story" to three
audience members whose ticket numbers he had drawn at random.

The Association's sincere thanks go to Bill Tate for organizing this
quarterly meeting and the talk by Captain Pearson -- it was largely
Bill's efforts that made the meeting the great success that it was.
